---
output: github_document

bibliography: references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

# IberianFishHSMs 

IberianFishHSMs: Compilation of models and tools to evaluate the habitat suitability at the microhabitat scale for Iberian freshwater fish species. The modelling approaches include sets of Habitat Suitability Curves (HSCs), Artificial Neural Networks (Multilayer Perceptrons, ANN-MLPs), Random Forests (RFs), Generalised Additive Models (GAMs), and Support Vector Machines (SVMs), all based on empirical data collected across several river basins in the Iberian Peninsula. The HSCs have been reviewed by an expert panel, which co-authored the package. The package also compiles sets of HSCs and Fuzzy Rule-Based Systems (FRBSs) based on expert knowledge, which was gathered through on-line forms. Functions for evaluating the habitat suitability from hydraulic simulations and for plotting the modelled suitability are also included. A thorough example of the capabilities of the packages appears below. The package includes 797 microhabitat suitability models for 43 species and up to 4 size classes (Large, Medium, Small, Very small) and spawning suitability (only for *Salmo trutta* and *Salmo salar*).

<br/>

## Authors

This package was developed by:

- **Rafael Muñoz-Mas**<sup>1</sup> (Author, Creator and Maintainer) - rafa.m.mas@gmail.com
- **Carlos Alonso**<sup>2</sup> (Contributor)
- **Enric Aparicio Manau**<sup>3</sup> (Contributor)
- **Fernando Cobo Gradín**<sup>4</sup> (Contributor)
- **Gustavo González Fernández**<sup>5</sup> (Contributor)
- **Javier Gortázar**<sup>6</sup> (Contributor)
- **Francisco Martínez-Capel**<sup>7</sup> (Contributor)
- **Francisco J. Oliva-Paterna**<sup>8</sup> (Contributor)
- **Jose Prenda**<sup>9</sup> (Contributor)
- **Jose Maria Santos**<sup>10</sup> (Contributor)
- **Rafael Miranda Ferreiro**<sup>11</sup> (Contributor)

<sup>1</sup> Tragsatec, Carrer Ramon Gordillo 7 - 46010 València, Spain.
<sup>2</sup>
<sup>3</sup> GRECO, Institute of Aquatic Ecology, University of Girona, E-17071 Girona.
<sup>4</sup> Laboratorio de Hidrobiologia, Departamento de Zoología, Genetica y Antropología Física. Facultad de Biología. Universidad de Santiago de Compostela, 15782 Santiago de Compostela, Spain.
<sup>5</sup> Icthios Gestión Ambiental SL. info@icthios.es.
<sup>6</sup> ECOHIDRÁULICA, S.L. C/ Villamanín 16, Madrid 28011.
<sup>7</sup> Institut d'Investigació per a la Gestió Integrada de Zones Costaneres (IGIC), Universitat Politècnica de València, C/Paranimf, 1, 46730 Gandia, València, Spain.
<sup>8</sup> Departamento de Zoología y Antropología Física. 30100. Universidad de Murcia. Murcia.
<sup>9</sup> Dpto. de Ciencias Integradas. Universidad de Huelva.
<sup>10</sup> Forest Research Centre, Associate Laboratory TERRA, School of Agriculture, University of Lisbon.
<sup>11</sup> Universidad de Navarra, Instituto de Biodiversidad y Medioambiente (BIOMA), 31008 Pamplona, Navarra, Spain.

<br/>

## Installation

IberianFishHSMs can be installed from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("RafaMMas/IberianFishHSMs")
```

<br/>

## Example

The following example illustrates the capabilities of IberianFishHSMs and the general workflow for predicting microhabitat suitability using outputs from hydraulic simulations and field data. Field data includes the substrate and cover (shelter) distribution across the study site, by patch, pixel or cell. The data used in this example is available from the package.

```{r load package, echo = TRUE}
library(IberianFishHSMs)
```

<br/>

### ListModels - List available models and their key characteristics.

`ListModels` allows inspecting the available models. The queries can be categorized by Species, Size, River, Model.type, Sampled.season, Valid.season, and/or Data.origin. This function also provides de Codes to internally call the microhabitat suitability models during the habitat evaluation. The output of this function also renders a summary of their main characteristics, including sample sizes, and several performance criteria for those models optimised through cross-validation.    

```{r ListModels, echo = TRUE}
Selected.models <- ListModels(Species = "Salmo trutta", verbose = F)

## General characteristics
Selected.models$Current.summary.table[,1:15]

## Performance

Selected.models$Current.summary.table[,16:23]

## Models and Codes

Selected.models[2:3]

```

`ListModels` output values can be use directly within `PredictHabitatSuitability` to carry out habitat evaluations. In particula, `$Codes` are the names used to store internally the available microhabitat suitability models.  

<br/>

### ListSelectedCoverTypes - List selected cover types in each microhabitat suitability model.

There exist an interaction between training data and the selected modelling technique, which can vary variable effects and importance [@Eugster2014]. By means of cross-validation, during the development of the microhabitat suitability models we carried out a variable selection approach discarding those cover types that not improved model performance for Machine Learning (i.e., ANN-MLPs, RFs, SVMs) and statistical approaches (i.e., GAMs) and an analogous approach during the development of the Habitat Suitability Curves/Criteria (HSCs) and Fuzzy Rule-based Systems (FRBSs).  This approach lead to different sets of relevant cover types in each model, although the core of relevant cover types is expected to coincide. The function `ListSelectedCoverTypes` allows inspecting the selected cover types for each model. The queries can be categorized by Species, Size, River, Model.type, Sampled.season, Valid.season, and/or Data.origin. When evaluating the microhabitat suitability, `PredictHabitatSuitability` selects and aggregates the appropriate cover types internally. Therefore, this function does not need to be called explicitly.

```{r ListSelectedCoverTypes, echo = TRUE}

ListSelectedCoverTypes(Species = "Salmo trutta")

```

<br/>

### PlotHabitatSuitabilityModels - Carry out sensitivity analyses

The package includes a number of models (i.e., 797) and species (i.e., 43) of several size classes and activities. Among other things, the available models vary by modelling technique, sampling site and the aggregation of data from different sites. The variable `Default` obtained with `ListModels` indicates which models are recommended because it was considered they performed and generalised better and/or were developed with data covering a large range of the microhabitat variables. Nevertheless, users can select other models when they consider it preferable. For example, when they are going to evaluate the microhabitat suitability in the basin where data for an alternative model was collected. In addition to `ListSelectedCoverTypes`, users can carry out sensitivity analyses to compare the predictions of the different available models. This allows inspecting their performance before selecting any alternative model or set of models. The sensitivity analysis can be carried out employing the function `PlotHabitatSuitabilityModels` as follows:

```{r, fig.width = 8, fig.height = 5, dpi=800}

## Pseudochondrostoma.polylepis, large, FRBS for Spring, Summer, and Autumn

(Selected.model <- ListModels(Species = "Pseudochondrostoma polylepis", verbose = FALSE)$Codes[1])

PlotHabitatSuitabilityModels(Selected.model = Selected.model, Quantiles = TRUE)

```

**Figure 1** - Sensitivity analysis (i.e., partial dependence plots) for *Pseudochondrostoma polylepis* large. The figure shows the results for the Fuzzy Rule-Based System (FRBS) to evaluate the microhabitat suitability during Spring, Summer, and Autumn.

This function generates partial dependence-like plots for microhabitat suitability models based on the provided data and selected model [@Friedman2001]. Partial dependence plots are created by varying one variable over its range of values while averaging the model's predictions over all the other features. This method isolates the effect of that single variable on the outcome and thus determining the overall effect. The function allows employing a user defined dataset to test especific combinations of variable values.

<br/>

### Habitat suitability prediction

Habitat suitability predictions are carried out based on mean flow velocity (m/s), water depth (m) substrate index (-) and the sum of the number of relevant types of cover in the evaluated microhabitats (i.e. pixels or cells from the hydraulic simulation). The package includes functions to prepare the input data. Additionally, the package includes data on a typical output from a hydraulic simulation and field survey. In particular, it includes data on velocity and depth for 40 simulated flows and the substrate composition and cover availability in the site of the Serpis River Basin (Eastern Iberian Peninsula) studied by Muñoz-Mas et all [-@Munoz-Mas2024].

``` {r Habitat assessment}
data("Velocity.example.df")
data("Depth.example.df")
data("Substrate.index.example.df")
data("Cover.example.df")

summary(Velocity.example.df)
summary(Depth.example.df)
summary(Substrate.index.example.df)
summary(Cover.example.df)

```

The function `SubstrateIndex` computes the substrate index as a weighted aggregation of the percentage of different granulometry classes (%). The substrate classes originally used corresponded to a simplification of the the American Geophysical Union size scale, namely silt (Ø ≤ 62 µm), sand (62 µm > Ø ≤ 2 mm), fine gravel (2 > Ø ≤ 8 mm), gravel (8 > Ø ≤ 64 mm), cobbles (64 > Ø ≤ 256 mm), boulders (Ø > 256 mm) and bedrock [see e.g., @Munoz-Mas2018 or @Munoz-Mas2017]. The package includes some examples to illustrate the structure of the input files.

```{r SubstrateIndex, echo = TRUE}
data(Substrate.index.example.df)

summary(Substrate.index.example.df)

head(SubstrateIndex(Substrate.index.example.df, check.completeness = FALSE)) # check.completeness allows inspecting whether names and percentages are correct.

Substrate.index <- SubstrateIndex(Substrate.index.example.df, check.completeness = FALSE)

```

```{r select model for evaluation}

Selected.species <- "Cobitis paludica"

Selected.size <- "Large"
  
(Selected.models <- ListModels(Species = Selected.species, Size = Selected.size)) ## example determining the selected models 
```

```{r}
Hydraulics <- data.frame(Velocity = Velocity.example.df$Velocity.0.05,
                         Depth = Depth.example.df$Depth.0.05,
                         Substrate.index = Substrate.index,
                         Cover.example.df[,-c(1:2)])

Predictions <- PredictHabitatSuitability(Selected.models = Selected.models, data = Hydraulics, HSC.aggregation = "geometric")

```

```{r}

head(Predictions)

```

```{r plot different predictions, fig.width = 14, fig.height = 14, dpi=800}
pairs(Predictions)
```

```{r plot and compare predicted microhabitat suitability, fig.width = 8, fig.height = 14, dpi=800}

library(sf)

Coordinates <- st_as_sf(Velocity.example.df[,c("x", "y")], coords = c("x", "y"))

Colors.suitability <- colorRampPalette(c("red2", "darkorange", "gold", "green2", "dodgerblue"))(100)

# sf_points <- st_as_sf(data.frame(Example.hydrulic.simulation.depth[,c("x", "y")], Predictions), coords = c("x", "y"), crs = 32630)

Model.names <- sapply(colnames(Predictions), function(x){
  x <- gsub(x = x, pattern = paste0(gsub(Selected.species, pattern = " ", replacement = "."), "."), replacement = "")
  x <- gsub(x = x, pattern = paste0(Selected.size, "."), replacement = "")
  x
})

op <- par(mfrow = c(ceiling(length(Selected.models$Models)^0.5), ceiling(length(Selected.models$Models)^0.5)), oma = c(0.5,0.5,2.5,0.5), mar = c(4,4,2,0.5))

for(i in 1:ncol(Predictions))
{
  plot(st_geometry(Coordinates), pch = 15, cex = 0.5, bty = "n", 
     col = Colors.suitability[cut(Predictions[,i], breaks = seq(0, 1, length = 101), labels = 1:100, include.lowest = T)], 
     xlab = "X", ylab = "Y")
  Axis(side = 1, cex.axis = 0.5)
  Axis(side = 2, cex.axis = 0.5) 
  title(Model.names[i], font.main = 1, cex.main = 0.9)
}

par(op)

Species <- Selected.species
Size <- paste0(" - ", Selected.size)
title(substitute(expr = paste(italic(Species), Size, sep =" "), env = list(Species = Species, Size = Size)), outer = T, adj = 0.05, cex.main = 2)

par(op)

```

<br/>

## Bibliography
