---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# IberianFishHSMs

<!-- badges: start -->
<!-- badges: end -->

IberianFishHSMs: Compilation of models and tools to evaluate the habitat suitability at the microhabitat scale for Iberian freshwater fish species. The modelling approaches include sets of Habitat Suitability Curves (HSCs), Artificial Neural Networks (Multilayer Perceptrons, ANN-MLPs), Random Forests (RFs), Generalised Additive Models (GAMs), and Support Vector Machines (SVMs), all based on empirical data collected across several river basins in the Iberian Peninsula. The HSCs have been reviewed by an expert panel which coauthored the package. The package also compiles sets of HSCs and Fuzzy Rule-Based Systems (FRBSs) based on expert knowledge, which was gathered through online forms. Functions for evaluating habitat suitability from hydraulic simulations and for plotting the modelled suitability are also included. An thorough example of the capabilitites of the packages appears below.

## Installation

IberianFishHSMs can be installed from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("RafaMMas/IberianFishHSMs")
```

## Example

IberianFishHSMs includes some examples to ilustrate the capabilities of the package and the general workflow to predict the habitat suitability from the outputs of hydraulic simulations and field surveys to colect data on substrate and cover (shelter) distribution across the study site.

```{r example}
library(IberianFishHSMs)
## basic example code
```

`ListModels` allows inpecting the vailable models. The queries can be categorized by Species, Size, River, Model.type, Sampled.season, Valid.season, and/or Data.origin. This function also provides de Codes to internally call the habitat suitability models during the habitat evaluation. The output of this function also renders a summary of their main characteristics, including sample sizes, and serveal perfomrance criteria.   

```{r cars}
ListModels(Species = "Salmo trutta")
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
